---
title: 'Exercise #3'
author: "Brian Schweigler; 16-102-071"
date: "23/03/2022"
output:
  pdf_document: default
  html_document: default
subtitle: t-test and R programming 
editor_options: 
  markdown: 
    wrap: 72
---

## Preliminaries
Loading the mean dataset:

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

mean20_df = read.csv("Mean20.txt", header = TRUE, comment.char = "#")
```

Handle the outliers:

-   Remove NAs

-   Remove illogical values (<= 0)


```{r}
temp_mean20_df <- mean20_df
temp_mean20_df[temp_mean20_df <= 0] <- NA
cleaned_mean20_df <- na.omit(temp_mean20_df)
summary(cleaned_mean20_df)
```
```{r}
sd(cleaned_mean20_df$time)
```
\newpage

## 2. Suppose that mean delay between two calls is 7.05 minutes. Can you test this hypothesis using the available data? What is your conclusion? Do you see a difference when considering the original values and the preprocessed values?


```{r}
assumed_mean <- 7.05
estimated_mean <- mean(cleaned_mean20_df$time)
t.test(x = cleaned_mean20_df, mu = assumed_mean, conf.level = 0.95)
```

Assuming we want the 95% confidance interval, we reject the hypothesis of the mean delay being 7.05 minutes.

Comparing to the unprocessed values:
```{r}
t.test(x = mean20_df, mu = assumed_mean, conf.level = 0.95)
```

While for the unprocessed data the true mean is not 7.05, it is clear that there is a >95% probability that our hypothesis holds. This is because the unprocessed data includes a negative value (I'd assume NA's are ignored), which increases the total range of values immensly.

\newpage

## 3. For Mary, the delay cannot be smaller than 7.05 minutes. Thus the only credible alternative hypothesis must take account of this (well-known) fact. How can you test Mary’s hypothesis?

We can use a one-sided t-test for this:

```{r}
t.test(x = cleaned_mean20_df, alternative = "g", mu = assumed_mean, conf.level = 0.95)
```
So for the processed data, this hypothesis holds in at least 95% of the cases.

```{r}
t.test(x = mean20_df, alternative = "g", mu = assumed_mean, conf.level = 0.95)
```
For the unprocessed variant this also holds true as the range of the 95% confidence interval is larger.

\newpage

## 4. Define a function secondMax(x), where x is a vector, returning the second largest value contained in x. If x is not a vector, return an error message. Test your implementation in different cases using the Mean20 dataset. 

First we define the function and check if x is a vector
```{r}
secondMax <- function(x) {
	if (!is.vector(x))	{
		return("Input x is not a numeric vector!")
	}
	# need at least 2 values
	if (length(x) <= 1) 	{
		return("x must contain at least 2 values!")
	}
	if (!is.numeric(x)) {
		return("Vector components must be numeric!")
	}
	
	# Remove largest value, such that the second largest value is now the largest
	x_without_max = x[-which(x == max(x))]
	
	return(max(x_without_max))
}

```


Testing the implementation:
```{r}
secondMax(cleaned_mean20_df$time)
secondMax(runif(n = 100, min = 0, max = 372)) # random values from 0 to 372
secondMax(c("1","2","4")) # should fail
secondMax(c("yes","this","is","patrick")) # should fail
secondMax(cleaned_mean20_df) # should fail
secondMax(NA) # should fail
secondMax(c(NA, NA, NA)) # should fail
```

\newpage

## 5. Define a function mySummary(x), where x is a vector composed by the mean, the median, the standard deviation, the minimum and the maximum values (in this order). Test your implementation in different cases using the Mean20 dataset 

```{r}
mySummary <- function(x, remove_na = TRUE) {
	if (!is.vector(x))	{
		return("Input x is not a numeric vector!")
	}
	# need at least 2 values
	if (length(x) <= 1) 	{
		return("x must contain at least 2 values!")
	}
	if (!is.numeric(x)) {
		return("Vector components must be numeric!")
	}
	if (remove_na) {
		na_values = which(is.na(x))
		if (length(na_values) > 0) {
			x <- x[-isna]
		}
	}
	
	return(c(mean = mean(x), median = median(x), stdev = sd(x), min = min(x), max = max(x)))
}
```


Testing with similar cases as before: 
```{r}
mySummary(cleaned_mean20_df$time)
mySummary(runif(n = 100, min = 0, max = 372)) # random values from 0 to 372
mySummary(c("1","2","4")) # should fail
mySummary(c("yes","this","is","patrick")) # should fail
mySummary(cleaned_mean20_df) # should fail
mySummary(NA) # should fail
mySummary(c(NA, NA, NA)) # should fail
```

