---
title: 'Exercise #4'
author: "Brian Schweigler; 16-102-071"
date: "06/04/2022"
output:
  pdf_document: default
  html_document: default
subtitle: Linear regression with R 
editor_options: 
  markdown: 
    wrap: 72
---

## Preliminaries
Loading the education dataset:

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

education_df = read.csv("EducationBis.txt", header = TRUE, sep = "\t", comment.char = "#")
```

Check for outliers:


```{r}
range(education_df$Wage)
```

```{r}
range(education_df$Education)
```

```{r}
unique(education_df$Gender)
```
No outliers directly visible, Gender is encoded as binary, years of education and wage range seem reasonable without additional information.

\newpage

## 1. Suppose that mean delay between two calls is 7.05 minutes. Can you test this hypothesis using the available data? What is your conclusion? Do you see a difference when considering the original values and the preprocessed values?


```{r}
assumed_mean <- 7.05
estimated_mean <- mean(cleaned_mean20_df$time)
t.test(x = cleaned_mean20_df, mu = assumed_mean, conf.level = 0.95)
```

Assuming we want the 95% confidance interval, we reject the hypothesis of the mean delay being 7.05 minutes.

Comparing to the unprocessed values:
```{r}
t.test(x = mean20_df, mu = assumed_mean, conf.level = 0.95)
```

While for the unprocessed data the true mean is not 7.05, it is clear that there is a >95% probability that our hypothesis holds. This is because the unprocessed data includes a negative value (I'd assume NA's are ignored), which increases the total range of values immensly.

\newpage

## 3. For Mary, the delay cannot be smaller than 7.05 minutes. Thus the only credible alternative hypothesis must take account of this (well-known) fact. How can you test Mary’s hypothesis?

We can use a one-sided t-test for this:

```{r}
t.test(x = cleaned_mean20_df, alternative = "g", mu = assumed_mean, conf.level = 0.95)
```
So for the processed data, this hypothesis holds in at least 95% of the cases.

```{r}
t.test(x = mean20_df, alternative = "g", mu = assumed_mean, conf.level = 0.95)
```
For the unprocessed variant this also holds true as the range of the 95% confidence interval is larger.

\newpage

## 4. Define a function secondMax(x), where x is a vector, returning the second largest value contained in x. If x is not a vector, return an error message. Test your implementation in different cases using the Mean20 dataset. 

First we define the function and check if x is a vector
```{r}
secondMax <- function(x)
{
	if (!is.vector(x))	{
		return("Input x is not a numeric vector!")
	}
	# need at least 2 values
	if (length(x) <= 1) 	{
		return("x must contain at least 2 values!")
	}
	if (!is.numeric(x)) {
		return("Vector components must be numeric!")
	}
	
	# Remove largest value, such that the second largest value is now the largest
	x_without_max <- x[-which(x == max(x))]
	
	return(max(x_without_max))
}

```


Testing the implementation:
```{r}
secondMax(cleaned_mean20_df$time)
secondMax(c("1","2","4")) # should fail
secondMax(c("yes","this","is","patrick")) # should fail
secondMax(cleaned_mean20_df) # should fail
secondMax(NA) # should fail
secondMax(c(NA, NA, NA)) # should fail
```

\newpage

## xyz

## Preliminaries
Loading the computers dataset:

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

computers_df = read.csv("Computers.txt", header = TRUE, sep =  , comment.char = "#")
```

Check for outliers:


```{r}
range(education_df$Wage)
```

```{r}
range(education_df$Education)
```

```{r}
unique(education_df$Gender)
```
No outliers directly visible, Gender is encoded as binary, years of education and wage range seem reasonable without additional information.

\newpage

## Preliminaries
Loading the cars dataset:

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

cars_df = read.csv("EducationBis.txt", header = TRUE, sep = " ", comment.char = "#")
```

Check for outliers:


```{r}
range(education_df$Wage)
```

```{r}
range(education_df$Education)
```

```{r}
unique(education_df$Gender)
```
No outliers directly visible, Gender is encoded as binary, years of education and wage range seem reasonable without additional information.

\newpage
